---
title: "Audit Trail Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Audit Trail Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyaudit)
library(dplyr)
```

## What is an audit trail?

When building data pipelines, it's easy to lose track of what happens at each
step. How many rows were dropped by that filter? Did the join introduce
duplicates? Which columns have missing values now?

tidyaudit's audit trail captures **metadata-only snapshots** at each step of a
pipe — row counts, column counts, NA totals, numeric summaries — without storing
the data itself. This gives you a lightweight, structured record of your
pipeline's behavior. The trail object also allows for custom functions to increase
flexibility and capture domain-specific diagnostics.

## Building a basic trail

Start by creating a trail object and inserting `audit_tap()` calls into your
pipeline. Each tap records a snapshot and passes the data through unchanged.

```{r basic-trail}
# Sample data
orders <- data.frame(
  id       = 1:20,
  customer = rep(c("Alice", "Bob", "Carol", "Dan", "Eve"), 4),
  amount   = c(150, 200, 50, 300, 75, 120, 400, 90, 250, 60,
               180, 210, 45, 320, 85, 130, 380, 95, 270, 55),
  status   = rep(c("complete", "pending", "complete", "cancelled", "complete"), 4)
)

trail <- audit_trail("order_pipeline")

result <- orders |>
  audit_tap(trail, "raw") |>
  filter(status == "complete") |>
  audit_tap(trail, "complete_only") |>
  mutate(tax = amount * 0.1) |>
  audit_tap(trail, "with_tax")
```

Now print the trail to see the snapshot timeline:

```{r print-trail}
print(trail)
```

The timeline shows row counts, column counts, NA totals, and change summaries
between consecutive steps. You can see exactly how many rows each filter removed
and when columns were added.

## Operation-aware taps

Plain `audit_tap()` records what the data looks like, but it can't tell you
*why* it changed. Operation-aware taps — `left_join_tap()`, `filter_tap()`,
etc. — perform the operation AND record enriched diagnostics.

### Join taps

Replace `dplyr::left_join()` + `audit_tap()` with `left_join_tap()` to capture
match rates, relationship type, and duplicate key information:

```{r join-tap}
customers <- data.frame(
  customer = c("Alice", "Bob", "Carol", "Dan"),
  region   = c("East", "West", "East", "North")
)

trail2 <- audit_trail("join_pipeline")

result2 <- orders |>
  audit_tap(trail2, "raw") |>
  left_join_tap(customers, by = "customer",
                .trail = trail2, .label = "with_region")

print(trail2)
```

The `Type` column now shows the join type, relationship, and match rate — all
without leaving the pipe.

All six dplyr join types are supported: `left_join_tap()`, `right_join_tap()`,
`inner_join_tap()`, `full_join_tap()`, `anti_join_tap()`, `semi_join_tap()`.

### Filter taps

`filter_tap()` keeps matching rows (like `dplyr::filter()`) while recording how
many rows were dropped:

```{r filter-tap}
trail3 <- audit_trail("filter_pipeline")

result3 <- orders |>
  audit_tap(trail3, "raw") |>
  filter_tap(status == "complete",
             .trail = trail3, .label = "complete_only") |>
  filter_tap(amount > 100,
             .trail = trail3, .label = "high_value",
             .stat = amount)

print(trail3)
```

The `.stat` argument tracks a numeric column through the filter, reporting how
much of the total was dropped — useful for financial pipelines where you want to
know the monetary impact of each filter.

`filter_out_tap()` works the same way but drops matching rows (the inverse).

## Comparing snapshots

`audit_diff()` gives you a detailed before/after comparison between any two
snapshots in the trail:

```{r audit-diff}
audit_diff(trail3, "raw", "high_value")
```

This shows row/column/NA deltas, columns added or removed, and numeric
distribution shifts.

## Full audit report

`audit_report()` prints the complete trail summary plus all consecutive diffs in
one call:

```{r audit-report}
audit_report(trail3)
```

## Tips

### Custom diagnostics

Pass a named list of functions via `.fns` to compute domain-specific diagnostics
at any tap. Each function receives the snapshot data as its argument (`.` in
formula lambdas) and its return value is stored in the snapshot.

```{r custom-fns}
trail4 <- audit_trail("custom_example")

result4 <- orders |>
  audit_tap(trail4, "raw", .fns = list(
    n_complete   = ~sum(.$status == "complete"),          # scalar
    amount_stats = ~c(mean = mean(.$amount),              # named vector
                      max  = max(.$amount))
  )) |>
  filter(status == "complete") |>
  audit_tap(trail4, "complete_only", .fns = list(
    n_complete   = ~sum(.$status == "complete"),
    amount_stats = ~c(mean = mean(.$amount),
                      max  = max(.$amount))
  ))
```

When you print the trail, custom results appear as inline `↳` annotations
directly below each snapshot row:

```{r print-custom}
print(trail4)
```

The rendering rules are:

- **Scalar** return value → `↳ fn_name: value`
- **Named vector or named list of scalars** → `↳ fn_name: key=val, key=val`
  (truncated at 60 characters)
- **Complex object** (data frame, unnamed vector, nested list) →
  `↳ fn_name: [complex — see audit_report()]`

To suppress annotations and display only the main table:

```{r show-custom-false}
print(trail4, show_custom = FALSE)
```

### NULL-trail mode

All tap functions work without a trail. When `.trail = NULL` (the default):

- **No diagnostic args**: behaves like the plain dplyr function
- **With `.stat` or `.warn_threshold`**: runs diagnostics and prints results
  without recording to a trail

```{r null-trail}
# Plain filter — no diagnostics
orders |> filter_tap(amount > 100) |> nrow()

# Diagnostics without a trail
orders |> filter_tap(amount > 100, .stat = amount) |> invisible()
```

This makes it easy to add quick diagnostics to any pipeline without setting up a
full trail.

## Snapshot controls

On wide datasets, computing numeric summaries for every column is unnecessary
and slows down the pipeline. Three parameters on all tap functions let you
control what gets captured:

- `.numeric_summary = FALSE` — skips quantile computation entirely.
- `.cols_include` — character vector of column names to include in the
  snapshot's schema (mutually exclusive with `.cols_exclude`).
- `.cols_exclude` — character vector of column names to exclude.

Core invariants — `nrow`, `ncol`, and `total_nas` — are always recorded
regardless of these settings.

```{r snapshot-controls}
wide_data <- cbind(orders, matrix(rnorm(20 * 50), nrow = 20))

trail_ctrl <- audit_trail("snapshot_controls")

wide_data |>
  audit_tap(trail_ctrl, "full_snapshot") |>
  audit_tap(trail_ctrl, "minimal",
            .numeric_summary = FALSE,
            .cols_include = c("id", "amount", "status"))

print(trail_ctrl)
```

The "minimal" snapshot still knows the data has `r ncol(wide_data)` columns and
20 rows, but its schema only describes the three columns you asked for and
contains no numeric summaries.

## Tabulation in pipelines

`tab()` produces one-way frequency tables or two-way crosstabulations:

```{r tab-standalone}
tab(orders, status)
tab(orders, status, customer)
```

`tab_tap()` embeds a tabulation inside a pipeline as a custom diagnostic
annotation. It runs `tab()` on the data, stores the result in the snapshot, and
returns `.data` unchanged — a transparent pass-through. This is useful for
tracking how distributions shift across pipeline steps.

```{r tab-tap}
trail_tab <- audit_trail("tab_pipeline")

result_tab <- orders |>
  tab_tap(status, .trail = trail_tab, .label = "status_dist") |>
  filter(status == "complete") |>
  tab_tap(customer, .trail = trail_tab, .label = "customer_dist",
          .sort = "freq_desc")

print(trail_tab)
```

## Exporting and serializing trails

Trails live in-memory as environment-based S3 objects. To share them with CI
systems, dashboards, or documentation you need to convert them to portable
formats. tidyaudit provides four serialization functions plus an HTML
visualization.

### Converting trails to R objects

```{r trail-to-objects}
# As a plain R list (suitable for jsonlite::toJSON())
trail_list <- trail_to_list(trail3)
str(trail_list, max.level = 2)

# As a data.frame (one row per snapshot)
trail_df <- trail_to_df(trail3)
print(trail_df)
```

### Saving and loading trails

RDS format preserves all R types and round-trips perfectly:

```{r trail-rds}
tmp_rds <- tempfile(fileext = ".rds")
write_trail(trail3, tmp_rds)
restored <- read_trail(tmp_rds)
print(restored)
```

JSON format is available for interoperability (requires jsonlite):

```{r trail-json}
tmp_json <- tempfile(fileext = ".json")
write_trail(trail3, tmp_json, format = "json")
```

### HTML visualization

`audit_export()` produces a self-contained HTML file — one file you can email
or embed in documentation:

```{r trail-html}
audit_export(trail3, tempfile(fileext = ".html"))
```

The HTML page renders an interactive pipeline flow diagram with light/dark theme
toggle. Nodes are clickable and expand to show column schema and diagnostics.
Edges are clickable and show the full diff between adjacent snapshots. No server
or internet connection is required to view the file.
